# New Data Modelling

Use new qPCR data (2026-03-19) to reproduce the HRM-ML pipeline.

{{< include global-settings.qmd >}}

## Data Preprocess

Read plate layout and transform labels to DNA quantities.

```{r}
library(dplyr)

plate = read.csv(xfun::magic_path("modeling-plate-labels.csv")) |>
  mutate(label_E = ifelse(label_E == 0, 1, ifelse(label_E == 16, 0, 1/2^label_E)),
         label_P = ifelse(label_P == 0, 1, ifelse(label_P == 16, 0, 1/2^label_P)))
```

Read the new QuantStudio data file and process melting curves.

```{r}
library(mcmodel)
library(stringr)

filename = "data-raw/modeling-qPCR/2026-3-19 222158.txt"
all = read_quantstudio(filename)

sample = plate |>
  mutate(cycle = 30, rep = 1)

mc_new = quantstudio2mc(all, primer = "V4", plate = sample) |>
  filterData(from = 80, to = 90) |>
  transformData(step = 0.1)
```

Convert to wide data frame and save.

```{r}
data_new = mc_tbl2wider(mc_new)

head(data_new)

write.csv(data_new, file = "data-clean/20260319.csv", row.names = FALSE)
```

## Random Forest Modelling

Filter to gradient matrix wells and log2-transform targets.

```{r}
mc_ml_data = read.csv("data-clean/20260319.csv") |>
  dplyr::filter(well_position %in% gradient_matrix_well) |>
  mutate(label_E = log2(label_E), label_P = log2(label_P))

data = initial_split(mc_ml_data)
data_train = training(data)
data_test = testing(data)
```

Model specifications.

```{r}
rf_spec = rand_forest(mode = "regression", trees = 1000) |>
  set_engine("ranger", importance = 'impurity', num.threads = 10)

plot_model_metric = function(predictions, model_metrics, truth_label, color){
  annotation = paste(model_metrics[['.metric']],
                     format(round(model_metrics[['.estimate']], 2), digits = 2),
                     sep = ": ",
                     collapse = "\n")
  ggplot(predictions, aes(.data[[truth_label]], `.pred`, color = I(color))) +
    geom_point(shape = 21) +
    geom_abline(slope = 1, linetype = 'dashed', color = color) +
    annotate("text", x = -Inf, y = Inf, color = I(color), label = annotation,
             hjust = -0.1, vjust = 1.1) +
    coord_equal() +
    xlim(c(-16,0)) + ylim(c(-16, 0)) +
    labs(x = expression(log[2]*Q[true]), y = expression(log[2]*Q[pred]))
}
```

Fit Random Forest for each species.

```{r}
#| fig-asp: 0.4
#| fig-width: 8
p_rf_predict = lapply(seq_along(strain_label), function(i){
  train = data_train |> select(matches(strain_label[[i]]), starts_with("T"))
  test = data_test  |> select(matches(strain_label[[i]]), starts_with("T"))
  rec = recipe(as.formula(paste(strain_label[[i]], '.', sep = '~')),
               data = train)

  fit = workflow() |>
    add_recipe(rec) |>
    add_model(rf_spec) |>
    fit(train)

  prediction = augment(fit, new_data = test)

  metric = prediction |>
    metrics(truth = strain_label[[i]], estimate = .pred)

  plot_model_metric(prediction, metric, strain_label[[i]], strain_color[[i]])
})

plot_grid(plotlist = p_rf_predict, ncol = 2, labels = "auto")

ggsave(filename = "figures/figure-new-data-rf.jpg")
```
