# New Data Single Species Modeling

Reproduce @fig-linear-modelling-of-single-species with the new qPCR data (2026-03-19).

{{< include global-settings.qmd >}}

## Data Preprocess

Read plate layout and transform labels to DNA quantities.

```{r}
library(dplyr)

plate = read.csv(xfun::magic_path("modeling-plate-labels.csv")) |>
  mutate(label_E = ifelse(label_E == 0, 1, ifelse(label_E == 16, 0, 1/2^label_E)),
         label_P = ifelse(label_P == 0, 1, ifelse(label_P == 16, 0, 1/2^label_P)))
```

Read the new QuantStudio data file and process melting curves.

```{r}
library(mcmodel)
library(stringr)

filename = "data-raw/modeling-qPCR/2026-3-19 222158.txt"
quantstudio_raw = read_quantstudio(filename)
```

## Amplification plot

```{r}
qs_amplification = get_quantstudio_amplication(quantstudio_raw)
```

```{r}
p_amp = lapply(seq_along(strain_label), function(i){
  df = qs_amplification  |>
    dplyr::filter(well_position %in% strain_single_well[[i]]) |>
    left_join(plate)

  ggplot(df, aes(cycle, delta_rn, group = well_position)) +
    geom_line(aes(color = .data[[strain_label[[i]]]])) +
    geom_hline(yintercept = 0.124, linetype = 'dashed', color = 'grey') +
    geom_text(x = -Inf, y = 0.124, label = 'Ct threshold',
              hjust = -.25, vjust = -0.5, color = 'grey') +
    scale_color_gradient(
      high = strain_color[[i]], low = "white", trans = 'log2',
      labels = trans_format("log2", label_number_auto()),
      breaks = 2^c(0, -5, -10, -15)
    ) +
    labs(color = expression(log[2]*Q), y = 'fluorescence') +
    theme(legend.position = "inside",
          legend.position.inside = c(0.18,0.62))
})
```

## Melting curve gradient

```{r}
mc = quantstudio2mc(quantstudio_raw, plate = plate, primer = "V4")
mc = filterData(mc, from = 80, to = 90) |>
  transformData(step = 0.1)
```

```{r}
plot_mc_single = function(mc, label_color, vline_color){
  tm = mc_get_tm(mc, npeaks = 1) |> pull(peak_position) |> median()

  df = mc2tbl(mc) |>
    select(-date, -primer) |>
    summarize(derivative = median(derivative),
              .by = all_of(c("temperature", label_color)))

  ggplot(df,
         aes(temperature, derivative,
             color = .data[[label_color]],
             group = .data[[label_color]])) +
    geom_line() +
    geom_vline(aes(xintercept = I(tm)),
               linetype = 'dashed',
               color = vline_color) +
    geom_text(x = tm, y = Inf, hjust = -0.1, vjust = 2,
              label = paste0("Tm = ", tm, "°C"),
              color = vline_color)  +
    scale_x_continuous(breaks = c(80, 85, 90)) +
    labs(x = "temperature (°C)",
         y = "fluorescence",
         color = expression(log[2]*Q)) +
    theme(legend.position = "inside",
          legend.position.inside = c(0.75,0.5))
}

p_mc = lapply(seq_along(strain_label), function(i){
  filterData(mc, well_position = strain_single_well[[i]]) |>
    plot_mc_single(strain_label[[i]], strain_color[[i]]) +
    scale_color_gradient(
      high = strain_color[[i]],
      low = "white",
      trans = 'log2',
      labels = trans_format("log2", label_number_auto()),
      breaks = 2^c(0, -5, -10, -15)
    )
})
```

## Peak calling

```{r}
mono_mc = lapply(seq_along(strain_label), function(i){
  x = filterData(mc, well_position = strain_single_well[[i]]) |>
    transformData(step = 0.1)
  df = mc2tbl(x) |>
    select(-date, -primer) |>
    nest(data = c(temperature, derivative)) |>
    pivot_longer(cols = starts_with("label_"),
                 names_to = "strain",
                 values_to = "quantity")
  peaks = lapply(df$data,
                 mcmodel:::detect_tm,
                 zero = "+",
                 npeaks = 1,
                 sortstr = TRUE,
                 threshold = 0) |>
    bind_rows()
  bind_cols(df, peaks)
}) |>
  bind_rows() |>
  dplyr::filter(quantity != 0)

mono_mc$peak_area = sapply(1:nrow(mono_mc), function(i){
  d = mono_mc$data[[i]] |>
    filter(temperature >= mono_mc$peak_start[[i]],
           temperature <= mono_mc$peak_end[[i]])
  pracma::trapz(d$temperature, d$derivative)
})

peak_cols = paste('peak',
                  c('height','area','start','position','end'),
                  sep = "_")
mono_mc = mono_mc |>
  dplyr::select(well_position, strain, quantity, matches(peak_cols))
```

## Ct model of single species

```{r}
qs_results = get_quantstudio_result(quantstudio_raw) |>
  dplyr::select(well_position, ct) |>
  dplyr::mutate(ct = as.numeric(ct)) |>
  left_join(plate) |>
  pivot_longer(cols = starts_with("label_"),
               names_to = "strain",
               values_to = "quantity") |>
  dplyr::filter(quantity != 0, !is.na(ct))

mono_ct = qs_results |>
  dplyr::filter(well_position %in% unlist(strain_single_well))
```

## Comparison of Ct and MC modelling

```{r}
plot_model_metric = function(model, test_data, color){
  predictions = augment(model, newdata = test_data)
  metrics = metric_set(rmse, rsq, mae)
  model_metrics = metrics(predictions,
                          truth = 'log2quantity',
                          estimate = .fitted)
  annotation = paste(model_metrics[['.metric']],
                     round(model_metrics[['.estimate']], digits = 2),
                     sep = ": ",
                     collapse = "\n")
  ggplot(predictions, aes(`log2quantity`, `.fitted`, color = I(color))) +
    geom_point(shape = 21) +
    geom_abline(slope = 1, linetype = 'dashed', color = color) +
    annotate("text", x = -Inf, y = Inf, label = annotation,
             color = I(color),
             hjust = -0.1, vjust = 1.1) +
    coord_equal() +
    xlim(c(-16,0)) + ylim(c(-16, 0)) +
    labs(x = expression(log[2]*Q[true]), y = expression(log[2]*Q[pred]))
}

p_model_ct = lapply(seq_along(strain_label), function(i){
  data = mono_ct |>
    dplyr::filter(well_position %in% strain_single_well[[i]]) |>
    mutate(log2quantity = log2(quantity))
  data_split_single = initial_split(data)
  data_train_single = training(data_split_single)
  data_test_single = testing(data_split_single)
  model = lm(log2quantity ~ ct, data_train_single)
  p = plot_model_metric(model, data_test_single,
                        strain_color[[i]]) + labs(subtitle = 'Ct model')
  return(p)
})

p_model_mc = lapply(seq_along(strain_label), function(i){
  data = mono_mc |>
    dplyr::filter(well_position %in% strain_single_well[[i]]) |>
    mutate(log2quantity = log2(quantity))
  data_split_single = initial_split(data)
  data_train_single = training(data_split_single)
  data_test_single = testing(data_split_single)
  model = lm(log2quantity ~ peak_area + peak_height + peak_start + peak_end,
             data_train_single)
  plot_model_metric(model, data_test_single, strain_color[[i]]) +
    labs(subtitle = 'MC model')
})
```

## Combined Results

```{r}
#| label: fig-new-data-linear-modelling-of-single-species
#| fig-width: 9.5
#| fig-cap: Linear modelling and prediction of single species abundance with threshold cycle (Ct) and melting curve features using new qPCR data (2026-03-19).
plot_grid(plotlist = c(p_amp, p_mc, p_model_ct, p_model_mc),
          align = 'hv',
          ncol = 4, nrow = 2, labels = 'auto')

ggsave("figures/figure-new-data-single-species.jpg")
```
